---
bibliography: mediation.bib
csl: apa.csl
---

```{r GlobOpt, include = FALSE}
fig.dir <- './Fig/2_single_cell_eqtl/'
dir.create(fig.dir, recursive = TRUE)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.path = fig.dir, fig.width = 4, fig.height = 4)
library(dplyr)
library(tidyr)
library(readr)
source('Util-Figure.R')

num.int <- function(x) format(x, big.mark = ',')
num.sci <- function(x, d=2) format(x, scientific = TRUE, digits = d)
num.round <- function(x, d=2) round(x, d)
fast.similarity <- function(...) suppressMessages(text2vec::sim2(...))
################################################################
```

```{r}
## Read eQTL data with all the significant SNPs (lodds > 0 in any)
ct <- c('bulk', 'ast', 'inh', 'ex', 'oli', 'mic', 'end', 'opc')
ct.names <- c('Brain_DLPFC', 'Astrocyte', 'Inhibitory_Neuron', 'Excitatory_Neuron', 'Oligodendrocyte', 'Microglia', 'Endothelial', 'Oligo_Progenitor')

eqtl.file <- './eqtl_results/rosmap_multivar_eqtl_combined.txt.gz'
eqtl.tab <- read_tsv(eqtl.file) %>%
    mutate(celltype = gsub(celltype, pattern = '_tf', replacement = '')) %>%
    mutate(celltype = factor(celltype, ct, ct.names)) %>%
    mutate(celltype = as.character(celltype))

## A list of protein coding genes
rosmap.genes <- read_tsv('rosmap_rows.txt.gz') %>%
    separate(col = tracking_id, into = c('ensg', 'remove'), sep = '[.]') %>%
    select(-remove)

coding.genes <- read_tsv('gencode.v26.coding.txt.gz',
                         col_names=c('chr', 'tss', 'tes', 'strand', 'ensg', 'hgnc'),
                         col_types='ciiccc_')
coding.ensg <- coding.genes %>% select(ensg) %>%
    separate(col=ensg, into=c('ensg','remove'), sep='[.]') %>%
    select(ensg) %>% unlist(use.names = FALSE)
################################################################
```

```{r GeneOntologyFunctions}
## Summary of enrichment within cluster
summary.enrichment.stat <- function(tab) {
    ## What can we control beyond the set size?

    ## calculate hypergeometric p-value
    ## phyper(q, m, n, k, lower.tail = TRUE, log.p = FALSE)
    ##    q: vector of quantiles representing the number of white balls
    ##       (drawn without replacement from an urn which contains both
    ##       black and white balls).
    ##    m: the number of white balls in the urn.
    ##    n: the number of black balls in the urn.
    ##    k: the number of balls drawn from the urn.
    ##    p: probability, it must be between 0 and 1.

    m <- max(tab$go.size)         ## white balls
    n <- max(tab$ntot) - m        ## black balls
    k <- max(tab$n.drawn)         ## balls drawn
    q <- length(unique(tab$gene)) ## whilte balls drawn

    pval <- phyper(q, m, n, k, lower.tail = FALSE)

    data.frame(n.white = m,
               n.drawn = k,
               n.overlap = q,
               pval = pval)
}

run.enrichment <- function(genes.tot, gene.tab, min.go = 10, max.go = 500){

    ensembl <- biomaRt::useEnsembl(biomart='ensembl',
                                   dataset = 'hsapiens_gene_ensembl',
                                   version = 79)

    go.attr <- c('ensembl_gene_id', 'go_id', 'name_1006', 'namespace_1003')

    go.bm <- biomaRt::getBM(filters = 'ensembl_gene_id', values = genes.tot,
                            attributes = go.attr,
                            mart = ensembl)

    go.tab <- go.bm %>% (function(tab) {
        names(tab) <- c('gene', 'go', 'go.name', 'go.onto');
        return(tab) }) %>%
            filter(nchar(go.name) > 0)

    go.size.tab <- go.tab %>%
        group_by(go) %>%
            summarize(go.size = n()) %>%
                as.data.frame()

    ntot <- go.tab %>% dplyr::select(gene) %>%
        unique() %>% nrow()

    gene.cluster <- gene.tab %>%
        filter(gene %in% unique(go.tab$gene)) %>%
            as.data.frame()
    
    cluster.size.tab <-
        gene.cluster %>%
            group_by(cluster) %>%
                summarize(n.drawn = n())

    ret <-
        gene.cluster %>%
            left_join(go.tab, by = 'gene') %>%
                left_join(go.size.tab, by = 'go') %>%
                    mutate(ntot = ntot) %>%
                        filter(go.size >= min.go, go.size <= max.go) %>%
                            left_join(cluster.size.tab, by = 'cluster') %>%
                                group_by(cluster, go, go.name, go.onto) %>%
                                    dplyr::do(summary.enrichment.stat(.)) %>%
                                        as.data.frame()
}
```

## Learning regulatory mechanisms in the brain tissues and 7 cell types

```{r}
################################################################
lodds.cutoff <- 0
pip.cutoff <- 1/(1 + exp(-lodds.cutoff))

heritable.bulk <- eqtl.tab %>%
    filter(celltype == 'Brain_DLPFC', lodds > lodds.cutoff) %>%
    select(med.id) %>% unique()

heritable.bulk.coding <- heritable.bulk %>%
    filter(med.id %in% coding.ensg)

nsnps.bulk <- 
    eqtl.tab %>% filter(celltype == 'Brain_DLPFC') %>%
    filter(lodds > lodds.cutoff) %>%
    group_by(med.id) %>%
    summarize(nsnp = length(unique(snp.loc)))
################################################################
```

### `r nrow(heritable.bulk) %>% num.int()` genes are heritable in the Brain dorsolateral prefrontal cortex (tissue-level eQTL discovery)

We estimate local genetic regulatory models, fitting sparse linear Gaussian regression models with the Bayesian spike-slab prior [@Mitchell1988-cf; @Carbonetto2012-bb], including all the common variants observed within independent linkage disequilibrium (LD) blocks [@Berisa2016-cm].  Of the `r nrow(rosmap.genes) %>% num.int()` genes measured in the ROSMAP RNA-seq data [@Mostafavi2018-ak; @Ng2017-xm], we identify that `r nrow(heritable.bulk) %>% num.int()` genes are strongly heritable with posterior inclusion probability (PIP) cutoff $>$ `r num.round(pip.cutoff)`; of them, `r nrow(heritable.bulk.coding) %>% num.int()` are protein coding genes according to the GENCODE version 26.
Per gene, we observe strong fixed effects on average `r mean(nsnps.bulk$nsnp) %>% num.round()` SNPs (SD = `r sd(nsnps.bulk$nsnp) %>% num.round()`) and nearly `r (100 * mean(nsnps.bulk$nsnp > 1)) %>% num.round()`% of genes are robustly associated with two or more SNPs.


```{r}
################################################################
tot.celltype.tab <- eqtl.tab %>%
    filter(lodds > lodds.cutoff) %>%
    filter(celltype != 'Brain_DLPFC') %>%
    select(med.id, chr, snp.loc) %>% unique() %>%
    mutate(snp = paste0(chr, ':', snp.loc)) %>%
    summarize(heritable.genes = length(unique(med.id)),
              num.eQTLs = length(unique(snp))) %>%
    as.data.frame()

active.celltype.tab <- 
    eqtl.tab %>%
    filter(lodds > lodds.cutoff, celltype != 'Brain_DLPFC') %>%
    select(med.id, celltype) %>% unique() %>%
    group_by(med.id) %>%
    summarize(active.celltypes = length(unique(celltype)))

count.celltype.tab <- eqtl.tab %>%
    filter(lodds > lodds.cutoff) %>%
    select(med.id, chr, snp.loc, celltype) %>% unique() %>%
    mutate(snp = paste0(chr, ':', snp.loc)) %>%
    group_by(celltype) %>%
    summarize(heritable.genes = length(unique(med.id)),
              num.eQTLs = length(unique(snp))) %>%
    arrange(desc(heritable.genes)) %>%
    as.data.frame()

.count.ct <- function(.ct) {
    count.celltype.tab %>% filter(celltype == .ct) %>%
        select(heritable.genes) %>%
            unlist()
}

active.celltype.summary <- active.celltype.tab %>%
    group_by(active.celltypes) %>%
    summarize(count = n())

single.celltype.genes <- active.celltype.summary %>%
    filter(active.celltypes == 1) %>%
    select(count) %>%
    unlist()

bulk.snps <- eqtl.tab %>%
    filter(lodds > lodds.cutoff, celltype == 'Brain_DLPFC') %>%
    select(chr, snp.loc) %>% unique()

celltype.snps <- eqtl.tab %>%
    filter(lodds > lodds.cutoff, celltype != 'Brain_DLPFC') %>%
    select(chr, snp.loc) %>% unique()

num.bulk.snps <- bulk.snps %>% nrow()
num.celltype.snps <- celltype.snps %>% nrow()
num.new.celltype.snps <- celltype.snps %>% anti_join(bulk.snps) %>% nrow()

bulk.genes <- eqtl.tab %>%
    filter(lodds > lodds.cutoff, celltype == 'Brain_DLPFC') %>%
    select(med.id) %>% unique()

celltype.genes <- eqtl.tab %>%
    filter(lodds > lodds.cutoff, celltype != 'Brain_DLPFC') %>%
    select(med.id) %>% unique()

num.bulk.genes <- bulk.genes %>% nrow()
num.celltype.genes <- celltype.genes %>% nrow()
num.new.celltype.genes <- celltype.genes %>% anti_join(bulk.genes) %>% nrow()
################################################################
```

### Our factored eQTL method identifies celltype of action in `r num.int(tot.celltype.tab$heritable.genes)` genes (celltype-level eQTL discovery)

We can deconvolve active cell types of the tissue-level eQTL effects on `r nrow(heritable.bulk) %>% num.int()` genes with our previous results on the estimated fraction of 7 brain cell types on the Brain DLPFC samples (He _et al._ in preparation; Mathys _et al._, in review).  The effect size deconvolution by our factored eQTL method is almost equivalent to sub-sampling of individuals.  Our deconvolution models identify clear celltype specificities on `r num.int(tot.celltype.tab$heritable.genes)` genes out of the total heritable ones.

```{r Celltype_Decompose, as.is = TRUE}
tab.celltype <- count.celltype.tab %>%
    mutate(heritable.genes = num.int(heritable.genes)) %>%
    mutate(num.eQTLs = num.int(num.eQTLs)) %>%
    flextable::regulartable()
tab.celltype <- flextable::bold(tab.celltype, part = 'header')
tab.celltype
```

```{r Fig_Num_Active_Celltype, fig.cap = '__Histogram of the number of active celltypes within a gene__', fig.width = 3, fig.height = 3}
################################################################
.aes <- aes(x = as.numeric(active.celltypes), y = count)
plt.active.celltype <-
    gg.plot(active.celltype.summary, .aes) +
    geom_bar(stat = 'identity') +
    scale_y_log10(breaks = active.celltype.summary$count) +
    xlab('Number of active cell-types')
print(plt.active.celltype)

gg.save(filename = paste0(fig.dir, '/Fig_Num_Active_Celltype.pdf'),
        plot = plt.active.celltype, width = 3, height = 3)
################################################################
```

```{r Fig_Celltype_Jaccard, fig.width = 4, fig.height = 5, fig.cap = '__Jaccard coefficients of genes sharing between celltypes__'}
library(Matrix)
library(methods)

jaccard.mat <- eqtl.tab %>%
    filter(lodds > lodds.cutoff) %>%
    select(med.id, celltype) %>%
    unique() %>%
    mutate(val = 1) %>%
    spread(key = celltype, value = val, fill = 0) %>%
    as.data.frame() %>%
    select(-med.id) %>%
    t() %>%
    (function(x) as(x, 'sparseMatrix')) %>%
    fast.similarity(method = 'jaccard') %>%
    as.matrix()

jaccard.tab <- jaccard.mat %>%
    as.data.frame() %>%
    (function(mat) { mat %>% mutate(row = rownames(mat)) }) %>%
    gather(key = 'col', value = 'jaccard', -row)

ct.order <- order.pair(jaccard.tab %>% rename(weight = jaccard))

celltype.jaccard.values <- jaccard.mat %>% as.data.frame() %>%
    (function(mat) { mat %>% mutate(row = rownames(mat)) }) %>%
    gather(key = 'col', value = 'jaccard', -row) %>%
    filter(row > col) %>% select(jaccard) %>% unlist()

jaccard.scale <- scale_fill_gradientn(colors = c('white', 'orange', 'red'),
                                      breaks = c(0, 0.1, 0.25, 0.5, 1),
                                      limits = c(0, 1),
                                      trans = 'sqrt')

plt.celltype.jaccard <-
    (jaccard.tab %>%
     mutate(row = factor(row, ct.order$row)) %>%
     mutate(col = factor(col, ct.order$col))) %>%
    gg.plot(aes(x = row, y = col, fill = jaccard)) +
    geom_tile(color = 'gray') +
    jaccard.scale +
    theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1),
          axis.title = element_blank(),
          legend.position = 'bottom',
          legend.key.height = unit(.5, 'lines'),
          legend.key.width = unit(2, 'lines'))

count.scale <- scale_y_continuous(trans = 'sqrt')

plt.celltype.count <- 
    (count.celltype.tab %>%
     mutate(celltype = factor(celltype, ct.order$row))) %>%
    gg.plot(aes(x = celltype, y = heritable.genes)) +
    geom_bar(stat = 'identity', fill = 'gray50') +
    geom_text(aes(label = heritable.genes - 10), size = 2) +
    ylab('# genes') +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank())

plt <- grid.vcat(list(plt.celltype.count, plt.celltype.jaccard),
                 heights = c(.5, 2.5))

gg.save(filename = paste0(fig.dir, '/Fig_Celltype_Jaccard.pdf'),
        plot = plt, width = 4, height = 5)
```

Since we applied sparse Bayesian prior probability on the genetic effects of factored eQTL models.  We found `r num.int(single.celltype.genes)` genes are only active in a single cell type (`r num.round(100 * (single.celltype.genes / tot.celltype.tab$heritable.genes))`%).  Interestingly we discovered that `r num.int(num.new.celltype.snps)` out of `r num.int(num.celltype.snps)` SNPs became newly associated with gene expressions when we applied celltype contexts.


### Gene ontology analysis

TODO: write up

```{r Run_GO_Enrichment_1}
temp.file <- '2_result_single_cell_eqtl.go_overall.RData'

if(!file.exists(temp.file)) {

    gene.tab <- eqtl.tab %>%
        rename(gene = med.id, cluster = celltype) %>%
            filter(lodds > lodds.cutoff) %>%
                select(gene, cluster) %>%
                    unique() %>%
                        filter(gene %in% coding.ensg)

    enrichment.overall.tab <-
        run.enrichment(genes.tot = coding.ensg,
                       gene.tab = gene.tab)

    save(list = c('enrichment.overall.tab'), file = temp.file)
} else {
    load(temp.file)
}
```

```{r Run_GO_Enrichment_2}
temp.file <- '2_result_single_cell_eqtl.go_heritable.RData'

if(!file.exists(temp.file)) {

    gene.tab <- eqtl.tab %>%
        rename(gene = med.id, cluster = celltype) %>%
            filter(lodds > lodds.cutoff, cluster != 'Brain_DLPFC') %>%
                select(gene, cluster) %>%
                    unique() %>%
                        filter(gene %in% coding.ensg)

    enrichment.heritable.tab <-
        run.enrichment(genes.tot = unlist(bulk.genes),
                       gene.tab = gene.tab)

    save(list = c('enrichment.heritable.tab'), file = temp.file)
} else {
    load(temp.file)
}
```


```{r GO_Func}
plot.go.tab <- function(.enrichment.tab, go.names) {

    .tab.sorted <- .enrichment.tab %>%
        filter(go.name %in% go.names) %>%
            mutate(coverage = n.overlap / n.white) %>%
                as.data.frame()

    .tab.sorted <- .tab.sorted %>%
        (function(tab) {
            .order <- tab %>%
                mutate(row = cluster, col = go.name, weight = -log10(pval)) %>%
                    select(row, col, weight) %>%
                        order.pair();
            tab %>% mutate(GO = factor(go.name, .order$cols),
                       cluster = factor(cluster, .order$rows))
        })

    .aes <- aes(y = GO, x = cluster, fill = -log10(pval), size = coverage)

    ret <- gg.plot(.tab.sorted) + geom_point(.aes, pch = 22) +
        scale_fill_gradientn(colors = c('white', 'red')) +
            scale_size_continuous(range = c(1, 4), limits = c(0, 1)) +
                scale_x_discrete(position = 'top')

    ret <- ret +
        theme(axis.text.x = element_text(angle = 45, vjust = 0, hjust = 0),
              axis.title = element_blank())

    return(ret)
}

take.top.go.tab <- function(.tab, n.top = 5) {
    .tab %>%
        group_by(cluster) %>%
            mutate(coverage = n.overlap / n.white) %>%
                top_n(n.top, 1 - pval) %>%
                    arrange(cluster, pval) %>%
                        as.data.frame()
}

make.go.flex.tab <- function(.tab) {
    ret <- .tab %>%
        mutate(p.value = num.sci(pval)) %>%
            rename(GO = go.name) %>%
                select(cluster, GO, n.overlap, coverage, p.value)

    ret <- ret %>%
    flextable::regulartable() %>%
    flextable::merge_v(j = c('cluster')) %>%
    flextable::bold(part = 'header') %>%
    flextable::align(align = 'center', part = 'header') %>%
    flextable::align(j = ~ n.overlap, align = 'right', part = 'body') %>%
    flextable::align(j = ~ coverage, align = 'right', part = 'body') %>%
    flextable::align(j = ~ p.value, align = 'right', part = 'body') %>%
    flextable::align(j = ~ cluster, align = 'left', part = 'body') %>%
    flextable::align(j = ~ GO, align = 'left', part = 'body') %>%
    flextable::autofit()

    return(ret)
}
```


```{r}
bp.tab <- enrichment.overall.tab %>%
    filter(go.onto == 'biological_process') %>%
    take.top.go.tab()

cc.tab <- enrichment.overall.tab %>%
    filter(go.onto == 'cellular_component') %>%
    take.top.go.tab()

mf.tab <- enrichment.overall.tab %>%
    filter(go.onto == 'molecular_function') %>%
    take.top.go.tab()
```




```{r Fig_GO_BP, fig.width = 6, fig.height = 6, fig.cap = '__Gene ontology enrichment analysis shows distinctive biological processes in different celltypes.__'}

plt.go.bp <- plot.go.tab(enrichment.overall.tab, bp.tab$go.name)
print(plt.go.bp)
gg.save(filename = paste0(fig.dir, '/Fig_GO_BP.pdf'), plot = plt.go.bp,
        width = 6, height = 6)
```



```{r Fig_GO_CC, fig.width = 8, fig.height = 6, fig.cap = '__Gene ontology enrichment analysis shows distinctive cellular components in different celltypes.__'}

plt.go.cc <- plot.go.tab(enrichment.overall.tab, cc.tab$go.name)
print(plt.go.cc)
gg.save(filename = paste0(fig.dir, '/Fig_GO_CC.pdf'), plot = plt.go.cc,
        width = 8, height = 6)
```



```{r Fig_GO_MF, fig.width = 8, fig.height = 6, fig.cap = '__Gene ontology enrichment analysis shows distinctive molecular functions in different celltypes.__'}

plt.go.mf <- plot.go.tab(enrichment.overall.tab, mf.tab$go.name)
print(plt.go.mf)
gg.save(filename = paste0(fig.dir, '/Fig_GO_MF.pdf'), plot = plt.go.mf,
        width = 8, height = 6)
```







# Tables


```{r Tab_GO_BP, results = 'asis'}
bp.flex.tab <- make.go.flex.tab(bp.tab)
bp.flex.tab
```
__Table.__ Biological process.


```{r Tab_GO_CC, results = 'asis'}
cc.flex.tab <- make.go.flex.tab(cc.tab)
cc.flex.tab
```
__Table.__ Cellular component.


```{r Tab_GO_MF, results = 'asis'}
mf.flex.tab <- make.go.flex.tab(mf.tab)
mf.flex.tab
```
__Table.__ Cellular component.





# References
